Differential invariant signatures for planar Lie group transformations with application to images

### 2021-08-25 paper/newerpaper.tex
Some updates

\date{\today}
\title{Differential invariant signatures for planar Lie group transformations of images}
\author{Richard Brown, Stephen Marsland, and Robert McLachlan}

\begin{document}
\maketitle
\begin{abstract}
Changes in viewpoint, such as camera motion, impose transformations on the objects in an image, and these transformations will differ for objects at different positions relative to the camera. Typically these transformations will be elements of planar Lie groups. While there have been significant advances in object recognition using machine learning in recent years, the question of how to recognise an object in an image as its appearance varies through camera motion and similar effects remains open. We demonstrate how differential invariant signatures can be derived for each of the transformation groups, and how the underlying invariances reflect the group-subgroup structure of the transformation groups.

@@

\section{Introduction}
The appearance of a fixed object in a set of images can vary markedly depending upon such variables as pose and position relative to the camera, even though the object itself does not change. This is even more marked in scenes that consist of multiple objects at different distances and orientations to the camera against a fixed background. The appearance of each individual object will then chnge in different ways as a constant camera motion or other such transformation is applied. Typically the individual transformations will appear as the action of an element of a planar Lie group, for example the affine, or projective groups, or an angle-preserving transformation from the conformal group, depending upon the camera model.

For image recognition it is then natural to seek representations of objects that do not change as these various transformation groups act. If the space of objects is denoted $M$ and the appearance transformations form a group $\mathcal{G}$, then mathematically one seeks a way to study objects in $M / \mathcal{G}$. One way to achieve this is by transforming all objects to match a chosen target object by registration (the transformation of the appearance of one shape to match another, typically by gradient descent on a matching function, see e.g.,~\citep{Modersitzki03}). For example, for the group of similarity transformations (rotation and reflection, translation, and scale), a Procrustes alignment~\citep{Kendall1989} will place all of the shapes into the reference frame of one template example. However, analysis on these quotient spaces is rarely simple (see e.g.,~\citet{RiemannianStats} for a basic primer), and if one is interested in the object, and not the transformation, then it necessitates a lot of computational work for relatively little benefit. The alternative approach, of showing examples of each object after applying sufficiently many transformations that the classifier can learn to recognise the object after any group transformation, inevitably scales extremely badly.

An alternative option is to seek $\mathcal{G}$-invariant functions on $M$. Mathematical invariants provide a useful way to identify sets of objects that are the same up to some equivalence relation, since the functions are constant for objects in the same equivalence class (orbits of the group). For images, rather than considering the image of the object, one constructs an invariant signature of some kind, which represents the object modulo the action of elements from the relevant group. This has been considered extensively by Olver and co-authors for the case of outlines of objects~\citep{Calabi1998,Hoff2013}. We extend this by considering the case of signatures for images rather than outlines of curves under a variety of transformation groups of relevance to image analysis.

@@

@@

In this paper our aim is develop the mathematical underpinnings of constructing differential invariant signatures for images under the action of planar Lie groups, and to summarise the impediments to a practical implementation, including the number of derivatives required. We demonstrate three distinct methods of computing differential invariants: by tensor contraction, using the theory of transvectants, and via the method of moving frames, applying the methods to a variety of the planar Lie groups, providing sample signature sets for the various groups we consider, together with an example of a sample signature surface based on a random transformation of a simple, smooth image.

The set of groups that we consider are shown as a lattice in Fig.~\ref{fig:lattice}, where the size of the groups decrease down the page, and the arrows denote the subgroup relationship.  There is an important relationship between a group and its subgroups for invariants, which is that invariants of a group are automatically invariants of its subgroups. This means that care is needed to distinguish between the two, with further invariants that are known not to be invariant in the supergroup included. The finite-dimensional groups in this lattice were chosen because they are the only planar Lie groups that include translation and have locally primitive algebras (that is, for any open subset of the plane, there is no one-dimensional foliation that is left invariant by the action of the corresponding group~\citep{Gonzalez-Lopez90}). This makes them the groups most directly relevant to image transformations. The infinite-dimensional examples are likewise those most likely to be of interest for image transformation, in particular the full diffeomorphism group, which is commonly used for image registration for medical images, see e.g.,~\citet{YounesBook}.

@@

\section{Preliminaries}

\subsection{Concepts and Notation}

An signature set is the image of $m$ invariants
$\mathcal{I}=(I_1,I_2,\dots,I_m)$. For simplicity of notation we will write
$\mathcal{I}(f) \subset \mathbb{R}^m$ with the understanding that each signature component is
actually acting on the jet space $J^d(f, \Omega, k)$. See Section~\ref{sec:T2} for a basic
example. A signature set is complete for $f$ if it determines $f$ up to
transformations from $\mathcal{G}$, i.e., if $\mathcal{I}(\tilde
f)(\tilde\Omega) = \mathcal{I}(f)(\Omega)$ implies that there exists a
transformation $\varphi\in \mathcal{G}$ such that $\tilde f = \varphi\cdot f$
and $\tilde \Omega = \varphi(\Omega)$.

%Groups we've done:
%translations, E(2), Sim(2), SA(2), A(2), PSL(2,C), PSL(3,R), Diff\_con, Diff\_vol, Diff
\subsection{Test image}

\subsection{Getting Started: Translations}
\label{sec:T2}

@@

%Groups we've done:
%translations, E(2), Sim(2), SA(2), A(2), PSL(2,C), PSL(3,R), Diff\_con, Diff\_vol, Diff
\section{Differential invariants through tensor contraction}
\label{sec:contraction}

@@

A moving frame is a mapping into a group $\mathcal{G}$ that provides a frame of reference along a manifold.  It was extended by~\citet{Cartan35} to the `method of moving frames'. If the group $\mathcal{G}$ is a Lie group then it has a homogeneous space and a frame of reference can be applied on the manifold structure of the group. Considering the actions of such a transformation group enables the identification of invariants to that group since it defines a $\mathcal{G}$-equivariant mapping. A (right-) moving frame is a smooth $\mathcal{G}$-equivariant map $\rho : \mathcal{M} \to \mathcal{G}$ such that $\rho (g \cdot z) = \rho(z) \cdot g^{-1}$, where $z \in J^d$ and where $\cdot$ denotes the group operation of $\mathcal{G}$. It is necessary~\citep{Olver2013} for the group to act freely. This means that the group action needs to be extended to the entire jet space of derivatives up to order $d$ by implicit differentiation of the group action, a process known as prolongation.

@@

\begin{enumerate}
\item Prolong the group action to the jet space of $d$-th order derivatives: $F: J^d \to \mathbb{R}$. 
\item Apply Cartan normalisation:
    \begin{enumerate}
    \item Choose a local cross-section to the group orbits, i.e., a ($d - \dim \mathcal{G}$)-dimensional submanifold $\mathcal{K}$ that intersects transversally at most once with each orbit
    \item $\mathcal{K}$ is specified by $\dim \mathcal{G}$ independent equations $Z_i (z) = c_i$, for $z \in J^d$,  and $Z_i$ scalar-valued functions and $c_i$ constants.
    \item The right-moving frame $g = \rho(z)$ is found by solving the normalisation questions $Z_i (g \cdot z) = c_i$ for the group parameters $g$ (in terms of $z$). In other words, the moving frame is the transformation back to the cross-section.
    \end{enumerate}
\item The differential invariants are not affected by the action of the group: $I(g \cdot z) = I(z) \: \forall z \in \mathrm{dom} I$, or equivalently, if it is constant on the orbits. 
\item Hence $F$ is made invariant by transforming it with a group action: $I(z) = F(g \cdot z) |_{g=\rho(z)}$.  
\end{enumerate}

>>>>

\date{\today}
\title{Differential invariant signatures for planar Lie group transformations with application to images}
\author{Richard Brown, Stephen Marsland, and Robert McLachlan}

\begin{document}
\maketitle
\begin{abstract}
The actions of different various Lie groups underlie the change of appearance of objects in images as the viewpoint changes, e.g., through camera motion. While there have been significant advances in object recognition using machine learning in recent years, the question of how to recognise an object in an image as its appearance varies through camera motion and similar effects remains open. We demonstrate how differential invariant signatures can be derived for each of the transformation groups, and how the underlying invariances reflect the group-subgroup structure of the relevant Lie groups.

@@

\section{Introduction}
The appearance of a fixed object in a set of images can vary markedly depending upon such variables as pose and position relative to the camera, even though the object itself does not change. This is even more marked in scenes that consist of multiple objects at different distances and orientations to the camera against a fixed background. The appearance of each individual object will then chnnge in different ways as a constant camera motion or other such transformation is applied. The individual transformations will appear as the action of an element of a planar Lie group, for example the affine, or projective groups, or an angle-preserving transformation from the conformal group, depending upon the camera model.

It is then natural to seek representations of objects that do not change as these various transformation groups act. If the space of objects is denoted $M$ and the transformations form a Lie group $\mathcal{G}$, then mathematically one seeks a way to study objects in $M / \mathcal{G}$.  One way to achieve this is to seek $\mathcal{G}$-invariant functions on $M$. Mathematical invariants provide a useful way to identify sets of objects that are the same up to some equivalence relation, since the functions are constant for objects in the same equivalence class (orbits of the group). For images, rather than considering the image of the object, one constructs an invariant signature of some kind, which represents the object modulo the action of elements from the relevant group. This has been considered extensively by Olver and co-authors for the case of outlines of objects~\citep{Calabi1998,Hoff2013}. We extend this by considering the case of signatures for images rather than outlines of curves under a variety of planar Lie transformation groups, particularly those of relevance to image analysis.

In this paper our aim is to follow the mathematical underpinnings of constructing differential invariant signatures for images under the action of (nested) planar Lie groups, and to summarise the impediments to a practical implementation, including the number of derivatives required. We demonstrate three distinct methods of computing differential invariants: by tensor contraction, using the theory of transvectants, and via the method of moving frames, applying the methods to a variety of the planar Lie groups, providing sample signature sets for the various groups we consider, together with an example of a sample signature surface based on a random transformation of a simple, smooth image.

%An alternative option is by transforming all objects to match a chosen target object by registration (the transformation of the appearance of one shape to match another, typically by gradient descent on a matching function, see e.g.,~\citep{Modersitzki03}). For example, for the group of similarity transformations (rotation and reflection, translation, and scale), a Procrustes alignment~\citep{Kendall1989} will place all of the shapes into the reference frame of one template example. However, analysis on these quotient spaces is rarely simple (see e.g.,~\citet{RiemannianStats} for a basic primer), and if one is interested in the object, and not the transformation, then it necessitates a lot of computational work for relatively little benefit. The alternative approach, of showing examples of each object after applying sufficiently many transformations that the classifier can learn to recognise the object after any group transformation, inevitably scales extremely badly.

\subsection{Invariant signatures of images}

@@

A signature set is the image of $m$ invariants
$\mathcal{I}=(I_1,I_2,\dots,I_m)$. For simplicity of notation we will write
$\mathcal{I}(f) \subset \mathbb{R}^m$ with the understanding that each signature component is
actually acting on the jet space $J^d(f, \Omega, k)$. See Section~\ref{sec:T2} for a basic
example. A signature set is complete for $f$ if it determines $f$ up to
transformations from $\mathcal{G}$, i.e., if $\mathcal{I}(\tilde
f)(\tilde\Omega) = \mathcal{I}(f)(\Omega)$ implies that there exists a
transformation $\varphi\in \mathcal{G}$ such that $\tilde f = \varphi\cdot f$
and $\tilde \Omega = \varphi(\Omega)$.

@@

\subsection{Relevant groups}

The set of groups that we consider are shown as a lattice in Fig.~\ref{fig:lattice}, where the size of the groups decrease down the page, and the arrows denote the subgroup relationship.  There is an important relationship between a group and its subgroups for invariants, which is that invariants of a group are automatically invariants of its subgroups. This means that care is needed to distinguish between the two, with further invariants that are known not to be invariant in the supergroup included. The finite-dimensional groups in this lattice were chosen because they are the only planar Lie groups that include translation and have locally primitive algebras (that is, for any open subset of the plane, there is no one-dimensional foliation that is left invariant by the action of the corresponding group~\citep{Gonzalez-Lopez90}). This makes them the groups most directly relevant to image transformations. The infinite-dimensional (pseudo-Lie) examples are likewise those most likely to be of interest for image transformation, in particular the full diffeomorphism group, which is commonly used for image registration for medical images, see e.g.,~\citet{YounesBook}.

@@

@@

%Groups we've done:
%translations, E(2), Sim(2), SA(2), A(2), PSL(2,C), PSL(3,R), Diff\_con, Diff\_vol, Diff

%Groups we've done:
%translations, E(2), Sim(2), SA(2), A(2), PSL(2,C), PSL(3,R), Diff\_con, Diff\_vol, Diff
\section{Differential Invariants By Direct Computation}

\subsection{Translations}
\label{sec:T2}

\subsection{Differential invariants through tensor contraction}
\label{sec:contraction}

@@

A moving frame is a mapping into a group $\mathcal{G}$ that provides a frame of reference along a manifold.  It was extended by~\citet{Cartan35} to the `method of moving frames'. The actions of a Lie group $\mathcal{G}$ enables the identification of invariants to that group through the definition of a $\mathcal{G}$-equivariant mapping. A (right-) moving frame is a smooth $\mathcal{G}$-equivariant map $\rho : \mathcal{M} \to \mathcal{G}$ such that $\rho (g \cdot z) = \rho(z) \cdot g^{-1}$, where $z \in J^d$ and where $\cdot$ denotes the group operation of $\mathcal{G}$. It is necessary~\citep{Olver2013} for the group to act freely. This means that the group action needs to be extended to the entire jet space of derivatives up to order $d$ by implicit differentiation of the group action, a process known as prolongation.

@@

\begin{enumerate}
\item Prolong the group action to the jet space of $d$-th order derivatives $J^d$.
\item Apply Cartan normalisation:
    \begin{enumerate}
    \item Choose a local cross-section to the group orbits, i.e., a ($d - \dim \mathcal{G}$)-dimensional submanifold $\mathcal{K}$ that intersects transversally at most once with each orbit
    \item $\mathcal{K}$ is specified by $\dim \mathcal{G}$ independent equations $Z_i (z) = c_i$, for $z \in J^d$,  and $Z_i$ scalar-valued functions and $c_i$ constants.
    \item The right-moving frame $g = \rho(z)$ is found by solving the normalisation questions $Z_i (g \cdot z) = c_i$ for the group parameters $g$ (in terms of $z$). In other words, the moving frame is the transformation back to the cross-section.
    \end{enumerate}
\item The differential invariants are not affected by the action of the group: $I(g \cdot z) = I(z) \: \forall z \in \mathrm{dom} I$, or equivalently, are constant on the orbits. 
\end{enumerate}
